Leadership across cultures and contexts: an exploratory analysis

### 2020-05-13 leadership_across_cultures_contexts.Rmd
Minor revisions in discussion and conclusion.

Anthropology, on the other hand, has been criticized for exoticizing and essentializing the "Other" as part of its colonial history [@said1979; @abu2008]. @abu2008 [p. 470], for instance, notes that

@@

Within benefit/cost dimensions, there was relatively more evidence for mating and territory benefits than costs for both leaders and followers, and equal levels of evidence for protection from harm vs. risk of harm. For leaders compared to followers, there were noticably higher odds ratios for mating, inclusive fitness, material resources, and social services. See Figure \@ref(fig:benefitcostratio).

@@

Shamans and other leaders with supernatural qualities were surprisingly common in the enthographic record. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of shamanism as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, which are a form of knowledge or expertise, and *Experienced/accomplished*). See Figure \@ref(fig:shamanism).

@@

We then fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, with random effects for author nested within culture, comparing these to intercept-only models, and selecting the `r nrow(feature_models_aic)` models whose AIC values were at least two points less than the intercept-only models (only Cultural conformity was dropped). As with our leader dimension models, we then dropped each factor variable in turn and compared the change in AIC. The large increases for group context and leader sex indicated that including these factors substantially improved fit. See Figure \@ref(fig:heatmapdrop1)B. We then compared the levels of each factor that improved fit for each feature variable. Female leaders were associated with social, reproductive, and material success, and male leaders with prosociality. Miltary groups were associated the Organize feature, religious groups with Prosociality and Competencies, and and kin groups with Strategizing and Prosocial competencies. See Figures S\@ref(fig:heatmapfeaturesex), and S\@ref(fig:heatmapfeaturegroups).

@@

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. There is a long history of leadership styles being described as based on either force and strength (i.e., dominance) or respect and special skills (i.e., prestige). Such dual-models are rooted in the coupling of ethology and animal dominance hierarchies with ethnography and cultural prestige systems [see @mead_sex_1935; @sahlins_poor_1963; @lewis_leaders_1974; @tiger_dominance_1970; @bernstein_dominance_1976; @barkow_prestige_1975]. Synthesizing the ethnographic and ethological evidence, @kracke_force_1978 linked "force and persuassion" to psychological-affect between leaders and followers in small-scale societies and Barkow [-@barkow_prestige_1975; -@barkow_prestige_1980] suggested the evolution of human prestige was linked to intra- and inter-sexual selection of culturally acquired competencies.

@@

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*. Shamans often provide valued public good, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017]. @ng06-054 describes shaman leaders among the Ojibwa:

@@

<!-- DISCUSSES SUCCESS FEATURE -->
**Leader success**. Leaders, probably as a consequence of their capacities to provide prosocial benefits (e.g., resolving conflicts and being generous) and/or impose social costs (e.g., punishing and aggressiveness), are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. There is a history within managerial studies among WEIRD populations of contrasting *leaders* from *managers*. @zaleznik_managers_1977, for example, suggested "managers" seek systematicity, organization, and stability, whereas "leaders" are willing to accept a lack of control and creatively navigate uncertainty to produce novel insight and appeal to the emotions of followers. Political anthropology reveals, however, in the absence of state-based or large-scale institutions, local leaders are often tasked with both managing various economic systems while also providing social services and interacting with followers on an interpersonal level [@cohen_comparative_1967]. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individuals behavior thereby incentivizing followership versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limitations of most extant psychological, sociological, and managerial data, many social scientists are now diversifying their samples to better understand *human* psychology, behavior, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diveristy within populations that are now often categorized as non-WEIRD. Furthermore, higher-order categories such as "subsistence strategy" are probably less relevant for understanding the complexity of behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record presents an invaluable opportunity to explore patterns of human cultural and behavioral diversity. Evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in comparative ethnographic methods.

@@

While there is evidence of a dominace style of leadership involving coercive authority, confidence and aggression, as well as a prestige style involving special knowledge, experience, and innovation, we suggest rethinking dichotomous models of leadership styles and focusing on inter-individual variation in specific leadership competencies. There is likely to be substantial overlap in physical, spiritual, cognitive, and other prosocial endowments and the degree to which a leader relies on any of these capacities will be shaped by their individual levels of such capacities, the capacities of ptential rivals, as well as group context, their functions, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [e.g., @cheng_dominance_2019; @price_evolution_2014], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. It seems evident that cultural forces and ideologies play important roles in shaping leadership styles and follower preferencesm including among WEIRD populations and in national political movements around the world [e.g., @mcdermott_leadership_2018; @inglehart_trump_2016].

>>>>

Anthropology, on the other hand, has been criticized for exoticizing and essentializing the "Other" as part of its colonial history [@said1979; @abu2008; @reyes-garcia_small-scale_2017]. @abu2008 [p. 470], for instance, notes that

@@

Within benefit/cost dimensions, there was relatively more evidence for mating and territory benefits than costs for both leaders and followers, and equal levels of evidence for protection from harm vs. risk of harm. For leaders compared to followers, there were noticeably higher odds ratios for mating, inclusive fitness, material resources, and social services. See Figure \@ref(fig:benefitcostratio).

@@

Shamans and other leaders with supernatural qualities were surprisingly common in the ethnographic record. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of shamanism as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, which are a form of knowledge or expertise, and *Experienced/accomplished*). See Figure \@ref(fig:shamanism).

@@

We then fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, with random effects for author nested within culture, comparing these to intercept-only models, and selecting the `r nrow(feature_models_aic)` models whose AIC values were at least two points less than the intercept-only models (only Cultural conformity was dropped). As with our leader dimension models, we then dropped each factor variable in turn and compared the change in AIC. The large increases for group context and leader sex indicated that including these factors substantially improved fit. See Figure \@ref(fig:heatmapdrop1)B. We then compared the levels of each factor that improved fit for each feature variable. Female leaders were associated with social, reproductive, and material success, and male leaders with prosociality. Military groups were associated the Organize feature, religious groups with Prosociality and Competencies, and and kin groups with Strategizing and Prosocial competencies. See Figures S\@ref(fig:heatmapfeaturesex), and S\@ref(fig:heatmapfeaturegroups).

@@

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. There is a long history of leadership styles being described as based on either force and strength (i.e., dominance) or respect and special skills (i.e., prestige). Such dual-models are rooted in the coupling of ethology and animal dominance hierarchies with ethnography and cultural prestige systems [see @mead_sex_1935; @sahlins_poor_1963; @lewis_leaders_1974; @tiger_dominance_1970; @bernstein_dominance_1976; @barkow_prestige_1975]. Synthesizing the ethnographic and ethological evidence, @kracke_force_1978 linked "force and persuasion" to psychological-affect between leaders and followers in small-scale societies and Barkow [-@barkow_prestige_1975; -@barkow_prestige_1980] suggested the evolution of human prestige was linked to intra- and inter-sexual selection of culturally acquired competencies.

@@

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*. Shamans often provide valued public goods, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate superhuman abilities [@singh_cultural_2017]. @ng06-054 describes shaman leaders among the Ojibwa:

@@

<!-- DISCUSSES SUCCESS FEATURE -->
**Leader success**. Leaders, probably as a consequence of their capacities to provide prosocial benefits (e.g., resolving conflicts and being generous) and/or impose social costs (e.g., punishing and aggressiveness), are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Several of these measures of success often co-occur. Among the Trobriand horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. There is a history within managerial studies among WEIRD populations of contrasting *leaders* from *managers*. @zaleznik_managers_1977, for example, suggested "managers" seek systematicity, organization, and stability, whereas "leaders" are willing to accept a lack of control and creatively navigate uncertainty to produce novel insight and appeal to the emotions of followers. Political anthropology reveals, however, in the absence of state-based or large-scale institutions, local leaders are often tasked with both managing various economic systems while also providing social services and interacting with followers on an interpersonal level [@cohen_comparative_1967; @fried_evolution_1967]. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individuals behavior thereby incentivizing followership versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limitations of most extant psychological, sociological, and managerial data, many social scientists are now diversifying their samples to better understand *human* psychology, behavior, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diversity within populations that are now often categorized as non-WEIRD. Furthermore, higher-order categories such as "subsistence strategy" are probably less relevant for understanding the complexity of behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record presents an invaluable opportunity to explore patterns of human cultural and behavioral diversity. Evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in comparative ethnographic methods.

@@

While there is evidence of a dominance style of leadership involving coercive authority, confidence and aggression, as well as a prestige style involving special knowledge, experience, and innovation, we suggest rethinking dichotomous models of leadership styles and focusing on inter-individual variation in specific leadership competencies. There is likely to be substantial overlap in and interaction between physical, spiritual, cognitive, and other prosocial competencies and the degree to which a leader relies on such competencies will be shaped by their individual levels in them, the competencies of potential rivals, as well as their functions and group context, follower preferences, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [e.g., @cheng_dominance_2019; @price_evolution_2014], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. It seems evident that cultural forces and ideologies play important roles in shaping leadership styles and follower preferences including among WEIRD populations and in national political movements around the world [e.g., @mcdermott_leadership_2018; @inglehart_trump_2016].
